Correlation tests

### 2025-04-21 analysis/manuscript.tex
Improve intro

Transformative agreements are a much-discussed library licensing model
for transitioning subscription-based journal publishing to full open
access. Although these agreements can vary considerably, they are mainly
aimed at hybrid journal bundles, allowing authors from participating
institutions to publish open access in these journals while providing
reading access to the entire portfolio. Library consortia and funders
have begun to coordinate how these agreements are negotiated, managed
and evaluated. At the same time, efforts are underway to improve
transparency through open data on contracts and payments. However, this
information is still only partially available, which limits large-scale
studies of the impact of transformative agreements on the transition of
hybrid journals to full open access. By comparing open metadata from
hoaddata, an openly available dataset on hybrid open access based on
metadata from Crossref, OpenAlex and the cOAlition S Journal Checker
Tool, with the established bibliometric databases Scopus and Web of
Science, this study aims to provide a robust estimate of the impact of
transformative agreements on hybrid open access. This comparative
approach informs bibliometric research and practice by highlighting the
strengths and weaknesses of each data source individually and
collectively.

Since their initial proposal by the Max Planck Digital Library (Schimmer
et al., 2015), transformative agreements have led to rapid growth in the
number of negotiated agreements and open access articles (Jahn, 2025;
McCabe \& Mueller-Langer, 2024; Rothfritz et al., 2024). Before, only a
small proportion of articles in hybrid journals were openly available
(Björk, 2012; Laakso \& Björk, 2016; Piwowar et al., 2018), with
publications from only a few European countries with dedicated funding
policies and the high-energy physics SCOAP\^{}3 consortium standing out
(Huang et al., 2020; Kohls \& Mele, 2018; Pinfield, 2015;
Robinson-Garcia et al., 2020). Related coordination activities, in
particular through the OA2020 and ESAC (Efficiency and Standards for
Article Charges) initiatives, have also led to improved workflows
between library consortia and publishers (Campbell et al., 2022;
Geschuhn \& Stone, 2017). However, critics have raised concerns about
these agreements, particularly regarding market concentration (Butler et
al., 2023; Shu \& Larivière, 2023), the failure of most hybrid journals
to convert to full open access (Kiley, 2024; Momeni et al., 2021),
reduced competition and incentives (McCabe \& Mueller-Langer, 2024;
Schmal, 2024), and the continuation of publication fees that widen gaps
between well-resourced and under-resourced institutions (Babini et al.,
2022; Ross-Hellauer et al., 2022)

While consortia evaluations confirm the growth of open access through
transformative agreements and acknowledge increasing coordination, they
reach mixed policy conclusions. For instance, the British Jisc
evaluation (Brayman et al., 2024) found that the consortium's
transformative agreements had a significant impact on national open
access growth but a limited effect on the global open access transition,
suggesting a need to reevaluate their strategy. Similar policy
reconsiderations emerged in Norway and Sweden (Holden et al., 2023;
Widding, 2024). Meanwhile, the funder initiative cOAlition S recommended
ending financial support for these agreements in 2024, but still
considers hybrid open access to be compliant with funders' open access
policies. Conversely, the German DEAL consortium extended agreements
with major publishers until 2028, while also outside of Europe, more
transformative agreements are being put in place (for instance in Latin
America, as described in Muñoz-Vélez et al. (2024)). Responding to
criticism, the Max Planck Digital Library recently outlined strategies
to evolve the original model, focusing on achieving fully open access
while controlling costs and addressing identified weaknesses (Dér, 2025)

Despite controversies, transformative agreements can benefit the
landscape of open research information as library consortia negotiate
for comprehensive open metadata, helping avoid data analytics from
becoming a commodity controlled by publishers (Aspesi \& Brand, 2020;
McCabe \& Mueller-Langer, 2024). For instance, ESAC recommends using
Crossref (Hendricks et al., 2020) to share metadata supporting workflows
including open access licenses and funding information (Geschuhn \&
Stone, 2017). Simultaneously, the bibliometric community pushes for more
open metadata through this publisher-driven DOI registration platform
and continuously monitors the progress (N. J. van Eck \& Waltman, 2022).

Many open scholarly data services that build on Crossref have benefited
from these developments. A prominent example is OpenAlex, which has
gained attention for its inclusivity, openness and comprehensive
coverage of publication metadata (Priem et al., 2022). In a recent
evaluation, Céspedes et al. (2025) reviewed the current evidence base on
the strengths and weaknesses of OpenAlex. They highlight both its
increasing use in bibliometric research and its adoption by research
organisations. Comparative studies have generally found OpenAlex to be a
reliable alternative to commercial databases (Alperin et al., 2024;
Culbert et al., 2025), although data quality issues occurred for key
metadata fields required for open access analysis, including evidence of
open access status itself (Jahn et al., 2023; Simard et al., 2024), as
well as authorships (Culbert et al., 2025), affiliations (Zhang et al.,
2024), author roles (Jahn, 2025) and document types used to distinguish
between different article types (Haupka et al., 2024). However, these
studies highlight the rapid improvements and new possibilities of
combining open metadata and the curation efforts of OpenAlex in
conjunction with the growing awareness of the importance of open
metadata, as demonstrated by the Barcelona Declaration on Open Research
Information.

Consequently, there is growing interest in using open bibliometric
databases such as Crossref and OpenAlex to assess the impact of
transformative agreements. Similar to established proprietary
bibliometric databases such as Scopus and Web of Science, these data
sources lack specific data points, particularly about open access
payments, which are typically not shared (Haustein et al., 2024).
However, bibliometric data can be enriched with open metadata about
journals and institutions involved in transformative agreements (Jahn,
2025). Jonge et al. (2025) demonstrated this potential by combining data
from the cOAlition S Journal Checker tool and OpenAlex, achieving 89\%
accuracy when validated against Dutch research information. While open
data approaches show promise, understanding how these results compare
when the same methodology is applied to proprietary databases such as
Web of Science and Scopus remains unexplored.

This study compares open and proprietary data sources for evaluating
transformative agreements. It adapts the approach from Jahn (2025),
which combines data from the cOAlition S Journal Checker tool, Crossref
and OpenAlex, and applies the same methods to Web of Science and Scopus
data. This comparative approach allows the evaluation of different
bibliometric data sources to monitor transformative agreements, while
assessing their respective strengths and limitations. Specifically, this
study compares the indexing of more than 13,000 hybrid journals across
these sources, and examines whether consistent results can be obtained
in terms of open access adoption between 2019 and 2023, analysed by
publishers and countries. By comparing results across these data
sources, this large-scale study provides a more robust understanding of
the impact of transformative agreements on the transition to full open
access. It also reveals the comparative strengths and weaknesses of
different bibliometric databases in measuring scholarly publishing's
shift toward openness.

@@

As shown in Figure \ref{fig:workflow}, the methodology involved
comparing hoaddata, an openly available collection of open research
information on hybrid open access, with the bibliometric databases Web
of Science and Scopus. This section introduces the initial data sources,
followed by a presentation of the necessary data processing steps to
obtain eligible articles enabled by transformative agreements using
author roles (first and corresponding) and harmonised affiliation data.

@@

hoaddata, developed and maintained by the author to support open access
monitoring and research (e.g. Jahn (2025)), is an R data package that
regularly collects information on hybrid open access uptake from
multiple openly available data sources (Jahn, 2024). It combines
article-level metadata from Crossref (Hendricks et al., 2020) and
affilation metadata from OpenAlex (Priem et al., 2022) with
transformative agreement information from the Transformative Agreement
Data dump used by the cOAlition S Journal Checker Tool (JCT)\footnote{\url{https://journalcheckertool.org/transformative-agreements/}},
which links journal and institutional data about participating research
organisations to agreements in the ESAC registry.

hoaddata followed good practices for computational reproducibility using
R (Marwick et al., 2018). The package, which includes data, code, a test
suite and documentation, is openly available on GitHub. To ensure
computational reproducibility while aggregating the data, a GitHub
Actions continuous integration and delivery (CI/CD) workflow handles
data retrieval from the SUB Göttingen's open scholarly data warehouse
based on Google BigQuery, which provides high-performant programmatic
access to monthly snapshots of Crossref and OpenAlex.\footnote{\url{https://subugoe.github.io/scholcomm_analytics/data.html}}
The workflow has run regularly to fetch updates from these data sources
since 2022. The package version used in this study is 0.3, containing
data from the Crossref 2024-08 dump provided to Crossref Metadata Plus
subscribers and the OpenAlex 2024-08-29 monthly dump. It covers
agreements collected between July 2021 to July 2024 from the JCT. This
version including the computation log is available on GitHub
(https://github.com/subugoe/hoaddata/releases/tag/v.0.3).

@@

Clarivate Analytics' Web of Science (WoS) is a well-established
proprietary bibliometric database consisting of several collections
(Birkle et al., 2020). Web of Science is a selective database with a
focus on base research (Stahlschmidt \& Stephen, 2022; Visser et al.,
2021). The collections considered in this study were the Science
Citation Index Expanded (SCIE), Social Sciences Citation Index (SSCI)
and Arts \& Humanities Citation Index (AHCI).

@@

Elsevier's Scopus, launched in 2004, is another widely used proprietary
bibliometric database for measuring research (Baas et al., 2020).
Similar to Web of Science, Scopus is selective with regard to the
journals it indexes. However, its journal coverage is much broader than
that of the Web of Science collections considered in this study, as it
indexes a wider range of applied research journals (Singh et al., 2021;
Stahlschmidt \& Stephen, 2022; Visser et al., 2021). With detailed
metadata about article types, open access status information derived
from Unpaywall, author roles, and disambiguated affiliations, Scopus
also contains important data to assess open access uptake, although no
direct information regarding transformative agreements was available at
the time of the study.

@@

Following Jahn (2025), the starting point was a unified dataset of
several safeguarded JCT snapshots\footnote{\url{https://github.com/njahn82/jct_data}}.
JCT journal data were enriched with ISSN variants linked to ISSN-L. To
identify hybrid journals, a comprehensive exclusion of fully open access
journals was performed using multiple journal lists including the
Directory of Open Access Journals (DOAJ). The resulting hybrid journal
data were made available via hoaddata and were used to independently
determine the publication volume for each database.

@@

Because neither Web of Science nor Scopus supported ROR-ID at the time
of data retrieval, the institution identifier used by the JCT, a
two-step matching process was implemented to harmonise the affiliation
data. First, 2,782,540 articles from 6,457 institutions with ROR-IDs in
the JCT data since 2017 (according to hoaddata) were processed to map
the first authors' ROR-IDs to corresponding proprietary affiliation
identifiers in Web of Science and Scopus using DOI matching. Then, an
algorithm selected the most frequent ROR ID and proprietary identifier
pairs to handle multiple affiliations and organisational hierarchy
differences.

This process linked 6,375 ROR-IDs to 4,894 Scopus Affiliation IDs and
6,034 ROR-IDs to 2,422 enhanced affiliation strings in Web of Science.
Quality evaluation through random sampling of 50 pairs revealed an error
rate of 22\% for Web of Science (11 mismatches) and 6\% for Scopus
(three mismatches). Upon inspection, these mismatches primarily occurred
with less-represented institutions having only a few publications,
introduced through multiple affiliations of single authors. The
difference between databases suggests that Scopus' affiliation control
aligns more closely with ROR than with Web of Science.

@@

\centering
\begin{tabular}[t]{lrrr}
\toprule
\textbf{} & \textbf{hoaddata*} & \textbf{Web of Science} & \textbf{Scopus}\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Hybrid journal coverage}}\\
\hspace{1em}Active journals & 12,890 & 8,655 & 11,888\\
\hspace{1em}Active journals (core) & 12,888 & 8,655 & 11,878\\
\hspace{1em}Active journals (core) with OA & 11,348 & 8,392 & 11,313\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Publication volume}}\\
\hspace{1em}Total published articles & 9,740,015 & 8,616,053 & 8,117,644\\
\hspace{1em}Original articles & 8,158,425 & 6,708,083 & 7,317,703\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Digital Object Identifier (DOI) coverage}}\\
\hspace{1em}Articles with DOI & 9,740,015 & 7,713,796 & 8,105,112\\
\hspace{1em}Original articles with DOI & 8,158,425 & 6,695,661 & 7,314,327\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Open Access (OA) metrics}}\\
\hspace{1em}OA articles & 998,699 & 1,112,758 & 974,099\\
\hspace{1em}Original OA articles & 969,817 & 1,019,784 & 922,578\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Original articles with affiliation data}}\\
\hspace{1em}First author articles & 7,242,542 & 6,294,855 & 7,232,017\\
\hspace{1em}Corresponding author articles & 5,534,207 & 6,291,441 & 6,898,487\\
\bottomrule
\multicolumn{4}{l}{\rule{0pt}{1em}\textsuperscript{*} Journal article metadata from Crossref, except affiliations from OpenAlex}\\
\end{tabular}

@@

While hoaddata only covered articles with a DOI, Scopus and Web of
Science publication indicators were calculated using their database
identifier. A subsequent comparison of DOI coverage shows that
non-original articles in Web of Science often lacked a DOI. This was
particularly the case for meeting abstracts, which are notably prevalent
in Health Sciences journals (Melero-Fuentes et al., 2025) and are not
indexed by Scopus (Donner, 2017). Open access indicators were aggregated
by DOI, because Unpaywall only collects information on open access
status for articles with DOI. A closer look at original articles with
affiliation data, in line with related research (Zhang et al., 2024),
reveals a lack of affiliation data in the case of OpenAlex, the
affiliation data source used by hoaddata, compared to Web of Science and
Scopus. In particular, only approximately two-thirds of the articles
examined provided corresponding author affiliations. The proportion for
first authors was 89\%. At the time of writing, OpenAlex disclosed
limited coverage of corresponding authorship data \footnote{\url{https://docs.openalex.org/api-entities/works/work-object/authorship-object\#is_corresponding}}.
Therefore, only first author data for hoaddata were considered in the
following analysis.

@@

TTidyverse tools (Wickham et al., 2019) for the R programming language
(R Core Team, 2024) were used throughout this largely automated data
collection and analysis process. Rank correlations were calculated using
the Hmisc package (Harrell Jr, 2003). Source code analysis, including
all queries used to obtain the data, is available on GitHub
(\url{https://github.com/njahn82/hoa_validation}). An interactive
supplement exploring correlations between the data sources examined by
country and publisher is available via HuggingFace Spaces:
\url{https://huggingface.co/spaces/najkoja/hoa_replication}.

@@

Figure \ref{fig-upset_coverage_results} presents the coverage of hybrid
journals included in transformative agreements, based on the
intersections of the examined data sources. The intersection sets of
journals and articles are visualised as an UpSet graph (Krassowski,
2020; Lex et al., 2014). The analysis included hybrid journals that
published at least one open access article between 2019 and 2023, based
on open access status information from each database. Only original
articles were included in the analysis.

@@

Journal coverage analysis revealed that 66\% (n = 7,970) of the hybrid
journals included in transformative agreements were indexed in all three
databases (Figure \ref{fig-upset_coverage_results}A). The second largest
set consisted of journals indexed in both hoaddata and Scopus,
comprising 21\% (n = 2,595) of hybrid journals. Notably, 6\% (n = 739)
of journals were exclusively contained in hoaddata, while another 6\% (n
= 748) were found only in Scopus. Of these, 354 were available in Web of
Science. Upon inspection, this group of hybrid journals exclusively
covered in proprietary data sources mainly represented hybrid journals
for which no open access evidence could be retrieved from Crossref, the
open access evidence source for hoaddata.

In terms of article coverage, Figure \ref{fig-upset_coverage_results}B
shows the total publication volume per combination in terms of DOI
availability. The largest set of hybrid journals covering all three data
sources also represented the largest number of articles. In total, these
journals recorded 6,289,687 articles, represented by the blue bars. They
recorded 94\% of the original articles with a DOI indexed in the Web of
Science sample, and 86\% in Scopus. Another 657,697 articles were
exclusive to the intersection between Scopus and hoaddata. Exclusively
in hoaddata were 177,110 articles, and exclusively in the proprietary
databases were 325,194 articles.

@@

Figure \ref{fig-upset_coverage_results_publisher} presents the coverage
of hybrid journals in transformative agreements across data sources from
2019 to 2023 focusing on publisher portfolios. The analysis highlights
the dominance of the three largest publishers, Elsevier, Springer
Nature, and Wiley, which collectively accounted for 46\% of the hybrid
journals and 62\% of the articles published during this five-year
period. In terms of article volume, Elsevier led with 2,435,478 articles
(34\% of the total) published across 1,941 hybrid journals (16\% of the
total). Springer Nature followed with 1,237,700 articles (17\%) in 2,294
hybrid journals, although it recorded the largest number of hybrid
journals (19\%). Wiley accounted for 814,349 articles (11\%) in 1,337
hybrid journals (11\%). The remaining 54 publishers collectively
accounted for 2,733,994 articles (38\%) in 6,442 hybrid journals (54\%).

@@

The three largest publishers, Elsevier, Springer Nature, and Wiley, were
best represented at the intersection of all three data sources
(hoaddata, Scopus, and Web of Science). Together, they comprised 4,347
hybrid journals (55\% of the intersectional set) and dominated article
coverage (n = 4,143,663; 66\%) as determined through shared DOIs. When
examining the publication volume per journal (Figure
\ref{fig-upset_coverage_results_publisher}C), Elsevier published, on
average, the largest journals, followed by Springer Nature and Wiley.

>>>>

Transformative agreements are a much-discussed library licensing model
for transitioning subscription-based journal publishing to full open
access. Although these agreements can vary considerably, they are mainly
aimed at hybrid journal bundles, enabling authors from participating
institutions to publish open access in these journals while providing
reading access to the entire portfolio (Borrego et al., 2021;
Hinchliffe, 2019). Following their initial proposal in 2015 (Schimmer et
al., 2015), the number of transformative agreements has grown
substantially, resulting in increased open access publishing in hybrid
journals (Jahn, 2025; McCabe \& Mueller-Langer, 2024; Rothfritz et al.,
2024). Prior to these agreements, only a few articles in hybrid journals
were openly accessible (Björk, 2012; Laakso \& Björk, 2016; Piwowar et
al., 2018), except for some European countries that had dedicated
funding policies and the SCOAP\(^3\) consortium for high-energy physics
journals (Huang et al., 2020; Kohls \& Mele, 2018; Pinfield, 2015;
Robinson-Garcia et al., 2020). Coordination efforts through initiatives
such as OA2020 and Efficiency and Standards for Article Charges (ESAC)
have further streamlined workflows between library consortia and
publishers (Campbell et al., 2022; Geschuhn \& Stone, 2017). By April
2025, the ESAC Transformative Agreement Registry, the primary
community-driven evidence source, listed approximately 1,300
agreements.\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}}

However, critics have raised concerns about these agreements,
particularly regarding perpetuating issues related to hybrid open access
such as charging twice for reading and publishing (``double dipping'')
(Asai, 2023), market concentration (Butler et al., 2023; Shu \&
Larivière, 2023), and the failure of most hybrid journals to convert to
full open access (Kiley, 2024; Momeni et al., 2021). Additional concerns
include reduced competition and incentives (McCabe \& Mueller-Langer,
2024; Schmal, 2024), and the continuation of publication fees that widen
gaps between well-resourced and under-resourced institutions (Babini et
al., 2022; Ross-Hellauer et al., 2022).

Consortia evaluations generally confirm increased open access through
transformative agreements and better coordination. However, they reached
mixed conclusions about the effectiveness. For instance, the British
Jisc evaluation (Brayman et al., 2024) found that the consortium's
transformative agreements had a significant impact on national open
access growth but a limited effect on the global open access transition,
suggesting a need to reevaluate its strategy. Similar policy
reconsiderations emerged in Norway and Sweden (Holden et al., 2023;
Widding, 2024). Meanwhile, the funder initiative cOAlition S recommended
ending financial support for these agreements in 2024, but still
considers hybrid open access to be compliant with funders' open access
policies. By contrast, the German DEAL consortium extended agreements
with major publishers until 2028, and countries in the Global South have
adopted transformative agreements (Muñoz-Vélez et al., 2024). Responding
to criticism, the Max Planck Digital Library recently outlined
strategies to evolve the original model, which continue to focus on
achieving fully open access while controlling costs and addressing
identified shortcomings (Dér, 2025).

These discussions highlight the need for robust data to evaluate
transformative agreements. Importantly, these agreements themselves can
help improve this evidence base, as researchers and libraries advocate
for including open metadata in negotiations to help avoid data analytics
from becoming a commodity controlled by publishers (Aspesi \& Brand,
2020; McCabe \& Mueller-Langer, 2024). For instance, ESAC recommends
using Crossref to share metadata supporting workflows including open
access licenses and funding information to support discovery and
monitoring (Geschuhn \& Stone, 2017). Simultaneously, the bibliometric
community pushes for more open metadata through this publisher-driven
DOI registration platform and continuously monitors progress (van Eck \&
Waltman, 2024).

This push for comprehensive metadata has yielded results. A growing
number of open scholarly data services that build on publisher-provided
metadata through Crossref have benefited from increasing metadata
coverage, although differences between publishers can be observed (van
Eck \& Waltman, 2024). OpenAlex stands out as a prominent example that
primarily retrieves updated records from Crossref and has gained
attention for its inclusivity, openness and comprehensive coverage
(Priem et al., 2022). Comparative studies confirm it as a suitable
alternative to proprietary bibliometric databases (Alperin et al., 2024;
Culbert et al., 2025). Notably, Céspedes et al. (2025) synthesised
evidence on both OpenAlex's strengths and its shortcomings for
bibliometric analysis. These include challenges with open access status
identification (Jahn et al., 2023; Simard et al., 2024), author
attribution (Culbert et al., 2025), institutional affiliations (Zhang et
al., 2024), and document type classification (Haupka et al., 2024).

While promising for general bibliometric research, applying open
metadata to transformative agreements presents distinct challenges due
to their targeted journal coverage and complex funding attribution. With
invoicing data largely unavailable, previous studies have relied on
publication fee pricing lists or expenditures, and first and/or
corresponding author affiliations as proxies to attribute funding to
institutions (Butler et al., 2023; Haustein et al., 2024; Zhang et al.,
2022). Recent work has extended this approach and attributed open access
articles to transformative agreements by using data from the cOAlition S
Journal Checker Tool, which links journals and participating
institutions to transformative agreements included in the ESAC registry
(Jahn, 2025; Jonge et al., 2025). Yet, no systematic comparison exists
of how open versus proprietary data sources perform when analysing the
same hybrid journals covered by transformative agreements to identify
comparative strengths and shortcomings of data sources.

The present study addresses this gap by comparing open and proprietary
data sources for evaluating transformative agreements. It adapts the
approach from Jahn (2025), which combines data from the cOAlition S
Journal Checker tool, Crossref as publication metadata source including
open access and OpenAlex as author affiliation data source, and applies
the same methods to the established databases Scopus and Web of Science.

This comparative approach examines different bibliometric data sources
through two primary research questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  How does the coverage of hybrid journals included in transformative
  agreements compare across Crossref, Scopus and Web of Science in terms
  of journal indexing and article representation?
\item
  To what extent do estimates of articles enabled by transformative
  agreements between 2019 and 2023 differ when using different
  affiliation data sources (OpenAlex, Scopus and Web of Science) and
  different authorship attribution methods (first versus corresponding
  authorship)?
\end{enumerate}

By analysing more than 13,000 hybrid journals across these sources, this
study provides insights into whether consistent results can be obtained
when measuring open access adoption across publishers and countries. The
findings enhance understanding of transformative agreements' impact on
the transition to full open access while revealing the comparative
strengths and limitations of different bibliometric databases in
measuring scholarly publishing's shift toward openness.

@@

As shown in Figure \ref{fig:workflow}, the methodology involved
comparing hoaddata, an openly available collection of open research
information on hybrid open access, with the bibliometric databases
Scopus and Web of Science. This section introduces the initial data
sources, followed by a presentation of the necessary data processing
steps to obtain eligible articles enabled by transformative agreements
using author roles (first and corresponding) and harmonised affiliation
data.

@@

hoaddata, developed and maintained by the author to support open access
monitoring and research (Achterberg \& Jahn, 2023; Jahn, 2025), is an R
data package that regularly collects information on hybrid open access
uptake from multiple openly available data sources (Jahn, 2024). It
combines article-level metadata from Crossref (Hendricks et al., 2020)
and affilation metadata from OpenAlex (Priem et al., 2022) with
transformative agreement information from the Transformative Agreement
Data dump used by the cOAlition S Journal Checker Tool (JCT)\footnote{\url{https://journalcheckertool.org/transformative-agreements/}},
which links journal and institutional data about participating research
organisations to agreements in the ESAC Transformative Agreement
Registry\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}}.

hoaddata follows good practices for computational reproducibility using
R (Marwick et al., 2018). The package, which includes data, code, a test
suite and documentation, is openly available on GitHub. To ensure
computational reproducibility while aggregating the data, a GitHub
Actions continuous integration and delivery (CI/CD) workflow handles
data retrieval from the SUB Göttingen's open scholarly data warehouse
based on Google BigQuery, which provides high-performant programmatic
access to monthly snapshots of Crossref and OpenAlex.\footnote{\url{https://subugoe.github.io/scholcomm_analytics/data.html}}
The workflow has run regularly to fetch updates from these data sources
since 2022. The package version used in this study is 0.3, containing
data from the Crossref 2024-08 dump provided to Crossref Metadata Plus
subscribers and the OpenAlex 2024-08-29 monthly dump. It covers
agreements collected between July 2021 to July 2024 from the JCT. This
version including the computation log is available on GitHub
(https://github.com/subugoe/hoaddata/releases/tag/v.0.3).

@@

Clarivate Analytics' Web of Science (WoS) is a well-established
proprietary bibliometric database consisting of several collections
(Birkle et al., 2020). Web of Science is selective regarding journal
indexing with a focus on base research (Stahlschmidt \& Stephen, 2022;
Visser et al., 2021). The collections considered in this study were the
Science Citation Index Expanded (SCIE), Social Sciences Citation Index
(SSCI) and Arts \& Humanities Citation Index (AHCI).

@@

Elsevier's Scopus, launched in 2004, is another widely used proprietary
bibliometric database for measuring research (Baas et al., 2020).
Similar to Web of Science, Scopus is selective, but its journal coverage
is much broader than that of the Web of Science collections considered
in this study, as it indexes a wider range of applied research journals
(Singh et al., 2021; Stahlschmidt \& Stephen, 2022; Visser et al.,
2021). With detailed metadata about article types, open access status
information derived from Unpaywall, author roles, and disambiguated
affiliations, Scopus also contains important data to assess open access
uptake, although no direct information regarding transformative
agreements was available at the time of the study.

@@

Following Jahn (2025), the starting point was a unified dataset of
several safeguarded JCT snapshots\footnote{\url{https://github.com/njahn82/jct_data}}.
JCT journal data were enriched with ISSN variants linked to ISSN-L
according to the ISSN agency. To identify hybrid journals, a
comprehensive exclusion of fully open access journals was performed
using multiple journal lists including the Directory of Open Access
Journals (DOAJ). The resulting hybrid journal data were made available
via hoaddata and were used to independently determine the publication
volume for each database.

@@

Because neither Web of Science nor Scopus supported ROR-ID at the time
of data retrieval, the institution identifier used by the JCT, a
two-step matching process was implemented to harmonise the affiliation
data. First, 2,782,540 articles from 6,457 institutions with ROR-IDs in
the JCT data since 2017 (according to hoaddata) were processed to map
the first authors' ROR-IDs to corresponding proprietary affiliation
identifiers in Web of Science and Scopus using DOI matching. Then, an
algorithm selected the most frequent ROR-ID and proprietary organisation
identifier pairs to handle multiple affiliations and organisational
hierarchy differences.

This process linked 6,375 ROR-IDs to 4,894 Scopus Affiliation IDs and
6,034 ROR-IDs to 2,422 enhanced affiliation strings in Web of Science.
Quality evaluation through random sampling of 50 pairs revealed an error
rate of 22\% for Web of Science (11 mismatches) and 6\% for Scopus
(three mismatches). Upon inspection, these mismatches primarily occurred
with less-represented institutions having only a few publications,
introduced through multiple affiliations of single authors. The
difference between databases suggests that Scopus's affiliation control
is more closely aligned with ROR than the Web of Science is with ROR.

@@

\centering
\begin{tabular}[t]{lrrr}
\toprule
\textbf{} & \textbf{hoaddata*} & \textbf{Web of Science} & \textbf{Scopus}\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Hybrid journal coverage}}\\
\hspace{1em}Active journals & 12,890 & 8,655 & 11,888\\
\hspace{1em}Active journals (original) & 12,888 & 8,655 & 11,878\\
\hspace{1em}Active journals (original) with OA & 11,348 & 8,392 & 11,313\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Publication volume}}\\
\hspace{1em}Total published articles & 9,740,015 & 8,616,053 & 8,117,644\\
\hspace{1em}Original articles & 8,158,425 & 6,708,083 & 7,317,703\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Digital Object Identifier (DOI) coverage}}\\
\hspace{1em}Articles with DOI & 9,740,015 & 7,713,796 & 8,105,112\\
\hspace{1em}Original articles with DOI & 8,158,425 & 6,695,661 & 7,314,327\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Open Access (OA) metrics}}\\
\hspace{1em}OA articles & 998,699 & 1,112,758 & 974,099\\
\hspace{1em}Original OA articles & 969,817 & 1,019,784 & 922,578\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Original articles with affiliation data}}\\
\hspace{1em}First author articles & 7,242,542 & 6,294,855 & 7,232,017\\
\hspace{1em}Corresponding author articles & 5,534,207 & 6,291,441 & 6,898,487\\
\bottomrule
\multicolumn{4}{l}{\rule{0pt}{1em}\textsuperscript{*} Journal article metadata from Crossref, except affiliations from OpenAlex}\\
\end{tabular}

@@

While hoaddata only covered articles with a DOI, Scopus and Web of
Science publication indicators were calculated using their database
identifier. A subsequent comparison of DOI coverage shows that
non-original articles in Web of Science often lacked a DOI. This was
particularly the case for meeting abstracts, which are notably prevalent
in Health Sciences journals (Melero-Fuentes et al., 2025) and are not
indexed by Scopus (Donner, 2017). Open access indicators were aggregated
by DOI, because Unpaywall only collects information on open access
status for articles in Crossref. A closer look at original articles with
affiliation data reveals a lack of data on affiliations and author roles
in the case of OpenAlex, the affiliation data source used by hoaddata,
compared to Web of Science and Scopus. In particular, only approximately
two-thirds of the articles examined provided corresponding author
affiliations. The proportion for first authors was 89\%. At the time of
writing, OpenAlex disclosed limited coverage of corresponding authorship
data \footnote{\url{https://docs.openalex.org/api-entities/works/work-object/authorship-object\#is_corresponding}}.
Therefore, only first author data for hoaddata were considered in the
following analysis.

@@

Tidyverse tools (Wickham et al., 2019) for the R programming language (R
Core Team, 2024) were used throughout this largely automated data
collection and analysis process. Rank correlations were calculated using
the Hmisc package (Harrell Jr, 2003). Source code analysis, including
all queries used to obtain the data, is available on GitHub
(\url{https://github.com/njahn82/hoa_validation}). An interactive
supplement exploring correlations between the data sources examined by
country and publisher is available via HuggingFace Spaces:
\url{https://huggingface.co/spaces/najkoja/hoa_replication}.

@@

Figure \ref{fig-upset_coverage_results} presents the coverage of hybrid
journals included in transformative agreements, based on the
intersections of the examined data sources. The intersection sets of
journals (\ref{fig-upset_coverage_results}A) and articles
(\ref{fig-upset_coverage_results}B= are visualised as an UpSet graph
(Krassowski, 2020; Lex et al., 2014). The x-axis represents the set
intersections in a matrix layout. The analysis included hybrid journals
that published at least one open access article between 2019 and 2023,
based on open access status information from each database. Only
original articles were included in the analysis.

@@

Journal coverage analysis revealed that 66\% (n = 7,970) of the hybrid
journals included in transformative agreements were indexed in all three
databases (Figure \ref{fig-upset_coverage_results}A). The second largest
set consisted of journals indexed in both hoaddata and Scopus,
comprising 21\% (n = 2,595) of hybrid journals. Notably, 6\% (n = 739)
of journals were exclusively contained in hoaddata, while another 6\% (n
= 748) were found only in Scopus. Of these, 354 were also available in
Web of Science. Upon inspection, the group of hybrid journals
exclusively covered in proprietary data sources mainly represented
hybrid journals for which no open access evidence could be retrieved
from Crossref, the open access evidence source for hoaddata.

In terms of article coverage, Figure \ref{fig-upset_coverage_results}B
shows the total publication volume per combination in terms of DOI
availability. The largest set of hybrid journals covering all three data
sources also represented the largest number of articles. In total, these
journals recorded 6,289,687 articles, represented by the blue bars. They
recorded 94\% of the original articles with a DOI indexed in Web of
Science, and 86\% in Scopus. Another 657,697 articles were exclusive to
the intersection between Scopus and hoaddata. Exclusively in hoaddata
were 177,110 articles, and exclusively in the proprietary databases were
325,194 articles.

@@

Figure \ref{fig-upset_coverage_results_publisher} presents the coverage
of hybrid journals in transformative agreements across data sources from
2019 to 2023 focusing on publisher portfolios. The analysis highlights
the dominance of the three largest publishers, Elsevier, Springer
Nature, and Wiley, which collectively accounted for 47\% of the hybrid
journals and 62\% of the articles published during this five-year
period. In terms of article volume, Elsevier led with 2,441,358 articles
(33\% of the total) published across 1,951 hybrid journals (16\% of the
total). Springer Nature followed with 1,247,578 articles (17\%) in 2,311
hybrid journals, although it recorded the largest number of hybrid
journals (19\%). Wiley accounted for 858,939 articles (12\%) in 1,382
hybrid journals (11\%). The remaining 54 publishers collectively
accounted for 2,749,847 articles (38\%) in 6,476 hybrid journals (53\%).

@@

The three largest publishers, Elsevier, Springer Nature, and Wiley, were
best represented at the intersection of all three data sources
(hoaddata, Scopus, and Web of Science). Together, they comprised 4,384
hybrid journals (55\% of the intersectional set) and dominated article
coverage (n = 4,174,315; 66\%) as determined through shared DOIs. When
examining the publication volume per journal (Figure
\ref{fig-upset_coverage_results_publisher}C), Elsevier published, on
average, the largest journals.

A comparison of publisher portfolios across different indexing sets
demonstrates that publishers were not represented uniformly. Notably,
Springer Nature exhibited 519 hybrid journals exclusively indexed in
both hoaddata and Scopus. This set included journals from the Chinese
Academy of Science, German-language medical journals, and Eastern
European publications including the \emph{Journal of Mathematical
Sciences}, which also publishes English-language translations of
Russian-language works. Additionally, this subset included titles with a
broader disciplinary focus, such as \emph{SN Computer Science}, and
newly launched hybrid journals such as \emph{Nature Computational
Science}, which started in 2021 and were indexed in Scopus but not yet
in Web of Science. The set also captures discontinued journals,
providing further insights into the dynamics of journal publishing.

Examining publisher portfolios not covered by hoaddata, but present in
Scopus or Web of Science, identified several publishers with missing
Creative Commons (CC) license information in Crossref. In particular,
Emerald represented 322 journals with 86,409 articles, AIP Publishing
accounted for 24 journals with 64,898 articles, and World Scientific
recorded 87 journals with 42,531 articles. In total, 9 publishers did
not share CC licenses with Crossref and were therefore not represented
in hoaddata.
